Trends in atomistic simulation software usage [\versionnumber]

### 2021-09-21 manuscript.tex
ASAP version

\documentclass[9pt,review]{livecoms}

\usepackage{lipsum} %
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{booktabs}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}
\addbibresource{atomistic.software.bl.bib}

\newcommand{\versionnumber}{1.3}  %
\newcommand{\githubrepository}{\url{https://github.com/ltalirz/livecoms-atomistic-software}}  %

@@

\title{Trends in atomistic simulation software usage [\versionnumber]}

\author[1,2,3*]{Leopold Talirz}
\author[4]{Luca M. Ghiringhelli}
\author[1,3]{Berend Smit}
\affil[1]{Laboratory of Molecular Simulation (LSMO),
    Institut des Sciences et Ingenierie Chimiques,
    Valais, \'Ecole Polytechnique F\'ed\'erale de Lausanne,
    CH-1951 Sion, Switzerland}
\affil[2]{Theory and Simulation of Materials (THEOS),
    Facult\'e des Sciences et Techniques de l'Ing\'enieur,
    \'Ecole Polytechnique F\'ed\'erale de Lausanne,
    CH-1015 Lausanne, Switzerland}
\affil[3]{National Centre for Computational Design and Discovery
of Novel Materials (MARVEL), \'Ecole Polytechnique F\'ed\'erale de Lausanne,
CH-1015 Lausanne, Switzerland}
\affil[4]{The NOMAD Laboratory at the Fritz Haber Institute of the Max Planck Society and Humboldt University, Berlin, Germany}

\corr{leopold.talirz@gmail.com}{LT}  %

\orcid{Leopold Talirz}{0000-0002-1524-5903}
\orcid{Luca M. Ghiringhelli}{0000-0001-5099-3029}
\orcid{Berend Smit}{0000-0003-4653-8562}

@@

\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubissue{<issue>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Day Month Year}
\dateaccepted{Day Month Year}

@@

\begin{abstract}
Driven by the unprecedented computational power available to scientific research, the use of computers in solid-state physics, chemistry and materials science has been on a continuous rise.
This review focuses on the software used for the simulation of matter at the atomic scale.
We provide a comprehensive overview of major codes in the field, and analyze how citations to these codes in the academic literature have evolved since 2010.
An interactive version of the underlying data set is available at \href{https://atomistic.software}{https://atomistic.software}.
\end{abstract}

@@

@@

Scientists today have unprecedented access to computational power.
This statement would be unremarkable, were it not for the extent to which computational power has exploded.
Figure~\ref{fig:top500} shows the performance ranking of the top 500 supercomputers in the world over the last decades, including the performance of the top-ranked machine, the machine at the bottom of the list, and the sum of all 500.
Remarkably, the least-squares fit to the sum (green line) corresponds to a growth rate of ${\sim}75\%$ per year, which translates to a performance increase of more than one \emph{million} times over the last 25 years.
Similar improvements in commodity hardware mean that many of 2020's laptop computers would have made the top 500 list of the early 2000s \cite{Lehtola2021}.

First versions of quantum-chemistry codes, such as Gaussian \cite{Hehre1970}, 
were already released in the 1970s,
followed by force-field codes, such as GROMOS \cite{VANGUNSTEREN1982}, 
and periodic density-functional theory (DFT) codes, such as CASTEP \cite{Payne1992a}, in the 1980s and 1990s.
In other words, many of these atomistic simulation engines have been around during this explosion of computational power, continuously evolving to take advantage of new algorithms, processor architectures, increasing parallelism and, more recently, dedicated accelerator hardware.
Over time, they have developed from instruments for specialists to proven and tested tools in the arsenal of practitioners in physics, chemistry, and materials science.

\begin{table*}
    \centering

\begin{tabular}{lllS[table-format=5.0]}
    \toprule
    Title                                                                 & Authors                    & Journal    & {\# cited} \\ \midrule
{Generalized gradient approximation made simple} & Perdew, Burke, Ernzerhof   & PRL (1996) & 108099  \\
{Development of the Colle-Salvetti correlation-energy formula … }       & Lee, Yang, Parr            & PRB (1988) & 77473   \\
{Efficient iterative schemes for ab initio total-energy calculations …} & Kresse, Furthmüller        & PRB (1996) & 58176   \\
{Projector augmented-wave method                                      } & Blöchl                     & PRB (1994) & 43455   \\
{Self-consistent equations including exchange and correlation …       } & Kohn, Sham                 & PR (1965)  & 42795   \\
{From ultrasoft pseudopotentials to the projector augmented-wave …    } & Kresse, Joubert            & PRB (1999) & 42485   \\
{Special points for Brillouin-zone Integrations                       } & Monkhorst, Pack            & PRB (1976) & 41232   \\
{Density-functional exchange-energy approximation with correct …      } & Becke                      & PRA (1988) & 41142   \\
{Inhomogeneous electron gas                                           } & Hohenberg, Kohn            & PRB (1964) & 35445   \\
{Ab initio molecular dynamics for liquid metals                       } & Kresse, Hafner             & PRB (1993) & 23192   \\
\end{tabular}

    \caption{
        Top ten most highly cited articles published by the American Physical Society,
        all of which deal with density functional theory and its practical application.
        Data collected from the Web of Science on June 16th, 2021.
    }
    \label{tab:aps-most-cited}
\end{table*}

Records of the pervasive use of these tools can be found in the scientific literature.
In a 2014 survey, van Noorden et al.\ found that 12 of the top 100 most cited papers of all time were on density-functional theory \cite{VanNoorden2014}.
As with other examples in van Noorden's list, the flood of citations are indicative of papers being cited by the many practitioners (here: of density-functional theory) rather than the few method developers.
If one focuses the analysis on articles published in physics journals, the footprint of density-functional theory grows even further:
For example, table~\ref{tab:aps-most-cited} shows the top ten most cited papers published by journals of the American Physical Society.
All of them are related to density-functional theory and its application.

\begin{table*}
    \centering
    \begin{tabular}{llllS[table-format=7.0]} \toprule
        Software & Version & Main language& {Contributors 2020} & {Lines of code} \\ \midrule
        NWChem \cite{Apra2020}&7.0.2&Fortran& 11 &5034139\textsuperscript{*} \\
        LAMMPS \cite{Plimpton1995}&27May2021&C++& 51 & 1039872  \\
        CP2K \cite{Kuhne2020a}&8.2&Fortran & 38 & 911367 \\
        ABINIT \cite{Gonze2020}&9.4.2&Fortran& 21 & 719048 \\
Quantum ESPRESSO \cite{Giannozzi2020}&6.7.0&Fortran& 43& 604666 \\
GROMACS \cite{Abraham2015}&2021.2&C++& 25 & 589108 \\
Psi4 \cite{Smith2020}&1.3.2&C++& 34 & 496858 \\
SIESTA \cite{Garcia2020}&5.0.0-alpha&Fortran& 4 & 411829 \\
Octopus \cite{Tancogne-Dejean2020}&10.5&Fortran& 10 & 297825 \\
OpenMM \cite{Eastman2017}&7.5.1&C++& 21 & 256452 \\
xtb \cite{Bannwarth2021}&6.4.1&Fortran& 18 & 124344 \\\hline
Linux kernel&5.13&C& >4000 \cite{LinuxHistory2020} & 20904410\\
    \end{tabular}\\
    \caption{
        Counting lines of code for 11 popular open-source atomistic simulation engines (and the Linux kernel for comparison), using the latest releases as of June 2021.
        Line counts are determined by cloc v1.6.0 \cite{Danial2021} and exclude blank lines, comments, and markup languages (detailed reports in the supporting information).
        Contributors for the year 2020 were determined by counting the number of different committers to the source code from January 1st 2020 to January 1st 2021 (numbers for the Linux kernel are from 2019).
        \textsuperscript{*} Roughly 3 million lines of code of NWChem are computer-generated.
    }
    \label{tab:lines-of-code}
\end{table*}

There used to be a time when it was commonplace for computational condensed-matter physicists and quantum chemists to write their own electronic-structure code, and many of the atomistic-simulation engines that are in broad use today have started this way.
Over the years, however, many of these engines have developed into complex software distributions.
Table \ref{tab:lines-of-code} shows counts for the lines of code in some of today's popular open-source simulation engines: they range from hundreds of thousands to millions of lines of code, typically written in Fortran or C++, with similar numbers being reported for commercial packages \cite{Krylov2015}.
While statistics like these are by no means accurate measures of code complexity (and developers follow different approaches to packaging and outsourcing of functionality to external libraries), they nevertheless suggest that many of these code bases are too large to be sustained by any single person.

This poses important questions for how to sustain these software projects going forward: questions of funding, business models, and software licenses.
Proponents of the open-source route argue that it democratizes research and education by removing barriers for both users and developers, and that science carried out with commercial software is harder to verify and reproduce \cite{Stodden2009,Gezelter2015,Lehtola2021}.
The open-source model can also be adopted irrespective of the size of a code's target user group, while commercial activity tends to require a minimal market size.
On the other hand, the strong focus on innovation and development often found in open-source scientific software can negatively impact usability and quality of documentation \cite{Swarts2019}.
Proponents of commercial licenses argue that making academic software accessible to a broader community is a technical task best left to professional software engineers, and that the resulting gain in scientific productivity can easily outweigh the license fees that pay for it \cite{Krylov2015}.
We note that open-source and commercial activity do not necessarily exclude each other: numerous software companies are built around an open-source core, and we start seeing first examples in atomistic modelling as well (e.g. Molcas/OpenMolcas \cite{Fdez.Galvan2019}).
Overall, it has been suggested that ``scientific publications are a more sound metric [of the scientific impact of software] than either the price of a product or whether its source code is available in the public domain'' \cite{Krylov2015}.
Providing a peek into this citation record is one of the reasons for creating the \atsoft collection.

The other reason is a practical one: 
When young scientists start their first research project in atomistic simulations, they often have no grasp of the extent of this software ecosystem, let alone of current trends in the field -- at least this is the personal experience of the authors of this review.
Software choices in research groups are therefore often informed by what other members of the group already use.
This makes sense: colleagues have vetted the code for the type of problems the research group is working on, and built up expertise around which of the many knobs to turn in order to find the sweet spot between efficiency and accuracy.

@@

@@

@@

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/overview}
    \caption{
        Overview table of atomistic simulation engines, sorted by how often they are referenced on Google Scholar during the previous year (here: 2020).
        A drop-down menu provides access to annual citation data reaching back to the year 2010.
    }
    \label{fig:overview}

@@

@@

\subsection{Trends}

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR BEST PRACTICES GUIDE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,review]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% Use the "ASAPversion' option following article acceptance to add the DOI and relevant dates to the document footer.
% Use the 'pubversion' option for adding the citation and publication information to the document footer, when the LiveCoMS issue is finalized.
% The 'bestpractices' option for indicates that this is a best practices guide.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{booktabs}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}
%\usepackage[
%backend=biber,%
%bibstyle=numeric-comp,%
%sorting=none%
%]{biblatex}
\addbibresource{atomistic.software.bl.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/ltalirz/livecoms-atomistic-software}}  %this should be the main github repository for this article

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Trends in atomistic simulation software usage [\versionnumber]}

\author[1,2,3*]{Leopold Talirz}
\author[4]{Luca M. Ghiringhelli}
\author[1,3]{Berend Smit}
%\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
%\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
%\author[2*]{Firstname Surname}
\affil[1]{Laboratory of Molecular Simulation (LSMO),
    Institut des Sciences et Ingenierie Chimiques,
    Valais, \'Ecole Polytechnique F\'ed\'erale de Lausanne,
    CH-1951 Sion, Switzerland}
\affil[2]{Theory and Simulation of Materials (THEOS),
    Facult\'e des Sciences et Techniques de l'Ing\'enieur,
    \'Ecole Polytechnique F\'ed\'erale de Lausanne,
    CH-1015 Lausanne, Switzerland}
\affil[3]{National Centre for Computational Design and Discovery
of Novel Materials (MARVEL), \'Ecole Polytechnique F\'ed\'erale de Lausanne,
CH-1015 Lausanne, Switzerland}
\affil[4]{The NOMAD Laboratory at the Fritz Haber Institute of the Max Planck Society and Humboldt University, Berlin, Germany}

\corr{leopold.talirz@gmail.com}{LT}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
%\corr{email2@example.com}{FS}

\orcid{Leopold Talirz}{0000-0002-1524-5903}
\orcid{Luca M. Ghiringhelli}{0000-0001-5099-3029}
\orcid{Berend Smit}{0000-0003-4653-8562}
%\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}

%\contrib[\authfn{1}]{These authors contributed equally to this work}
%\contrib[\authfn{2}]{These authors also contributed equally to this work}

%\presentadd[\authfn{3}]{Department, Institute, Country}
%\presentadd[\authfn{4}]{Department, Institute, Country}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubissue{<issue>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Day Month Year}
\dateaccepted{Day Month Year}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% 250 words

\begin{abstract}
%(to be adapted to match the actual content of the article)
Driven by the unprecedented computational power available to scientific research, the use of computers in solid-state physics, chemistry and materials science has been on a continuous rise.
This review focuses on the software used for the simulation of matter at the atomic scale.
We provide a comprehensive overview of major codes in the field, and analyze how citations to these codes in the academic literature have evolved since 2010.
An interactive version of the underlying data set is available at \href{https://atomistic.software}{https://atomistic.software}.
\end{abstract}

@@

%\begin{itemize}
%    \item the use of computational methods in physics/chemistry/mat sci is on the rise
%    \item the complexity of the computationally feasible methods is on the rise
%    \item the complexity of the software implementing those methods is on the rise
%        %(the gradual decline of single-thread performance scaling is leading to increasing specialization of hardware whose features need to be exploited for high performance.
%        %this increases complexity and is a problem for small codes and may lead to a consolidation in the atomistic software space)
%    \item this poses important questions for how to sustain scientific software development going forward;
%        questions of business models, licensing, source availability, etc.
%    \item we are blessed with a lot of great open-source software in our field.
%        how do they fare with respect to their commercial competitors?
%    \item there are hundreds of simulation engines out there - difficult to keep track without getting lost
%    \item figures: 
%    \begin{itemize}
%        \item one figure on citation of atomistic codes (potentially just DFT) in all of science + the physics/chemistry subfields
%        \item word cloud of major atomistic simulation codes
%    \end{itemize}
%\end{itemize}

@@

% latest
% sum last 20 years (data points): 2.4e18/88.1e12=27.24k
% sum fit: roughly 270/10 years => 75% growth per year => 
% trend:

Scientists today have unprecedented access to computational power.
This statement would be unremarkable, were it not for the extent to which computational power has exploded.
Figure~\ref{fig:top500} shows the performance ranking of the top 500 supercomputers in the world over the last decades, including the performance of the top-ranked machine, the machine at the bottom of the list, and the sum of all 500.
Remarkably, the least-squares fit to the sum (green line) corresponds to a growth rate of ${\sim}75\%$ per year, which translates to a performance increase of more than one \emph{million} times over the last 25 years.
Similar improvements in commodity hardware mean that many of 2020's laptop computers would have made the top 500 list of the early 2000s \cite{Lehtola2021}.
%\footnote{High-end laptop processors exceed 100 GFlop/s in the Linpack benchmark.}

First versions of quantum-chemistry codes, such as Gaussian \cite{Hehre1970}, 
%or GAMESS, 
were already released in the 1970s,
followed by force-field codes, such as GROMOS \cite{vanGunsteren1982}, 
and periodic density-functional theory (DFT) codes, such as CASTEP \cite{Payne1992a}, in the 1980s and 1990s.
In other words, many of these atomistic simulation engines have been around during this explosion of computational power, continuously evolving to take advantage of new algorithms, processor architectures, increasing parallelism and, more recently, dedicated accelerator hardware.
Over time, they have developed from instruments for specialists to proven and tested tools in the arsenal of practitioners in physics, chemistry, and materials science.

%making previously complicated calculations routine
%and previously impossible calculations feasible.

% Gaussian with its first release in 1970,
%GAMESS in the lat 1970s.
%CASTEP in the late 1980s
%Turbomole 1987
%Dmol3 (early 80s)

\begin{table*}
    \centering
%\small

\begin{tabular}{lllS[table-format=5.0]}
    \toprule
    Title                                                                 & Authors                    & Journal    & {\# cited} \\ \midrule
{Generalized gradient approximation made simple} & Perdew, Burke, Ernzerhof   & PRL (1996) & 108099  \\
{Development of the Colle-Salvetti correlation-energy formula … }       & Lee, Yang, Parr            & PRB (1988) & 77473   \\
{Efficient iterative schemes for ab initio total-energy calculations …} & Kresse, Furthmüller        & PRB (1996) & 58176   \\
{Projector augmented-wave method                                      } & Blöchl                     & PRB (1994) & 43455   \\
{Self-consistent equations including exchange and correlation …       } & Kohn, Sham                 & PR (1965)  & 42795   \\
{From ultrasoft pseudopotentials to the projector augmented-wave …    } & Kresse, Joubert            & PRB (1999) & 42485   \\
{Special points for Brillouin-zone Integrations                       } & Monkhorst, Pack            & PRB (1976) & 41232   \\
{Density-functional exchange-energy approximation with correct …      } & Becke                      & PRA (1988) & 41142   \\
{Inhomogeneous electron gas                                           } & Hohenberg, Kohn            & PRB (1964) & 35445   \\
{Ab initio molecular dynamics for liquid metals                       } & Kresse, Hafner             & PRB (1993) & 23192   \\
%\red{Accurate and simple analytic representation of the electron-gas …    } & Perdew, Wang               & PRB (1992) & 19442   \\
%\red{Soft self-consistent pseudopotentials in a generalized eigenvalue …  } & Vanderbilt                 & PRB (1990) & 17617   \\
%\red{Atoms, molecules, solids, and surfaces - Application of the ...      } & Perdew, Chevary et al.     & PRB (1992) & 17066   \\
%{The electronic properties of graphene                                } & Castro Neto, Guinea et al. & RMP (2009) & 16687   \\
%\red{Self-interaction correction to density-functional approximations ... } & Perdew, Zunger             & PRB (1981) & 16259   \\
%\red{Density-functional approximation for the correlation-energy of ...   } & Perdew                     & PRB (1986) & 15390   \\
%{Fast algorithm for detecting community structure in networks         } & Newman                     & PRE (2004) & 14895   \\
%{Finding and evaluating community structure in networks               } & Newman, Girvan             & PRE (2004) & 14895   \\
%{Optical constants of noble metals                                    } & Johnson, Christy           & PRB (1972) & 14179   \\
%\red{Note on an approximation treatment for many-electron systems         } & Møller, Plesset            & PR (1934)  & 13320 \\ \bottomrule
\end{tabular}

    \caption{
        Top ten most highly cited articles published by the American Physical Society,
        all of which deal with density functional theory and its practical application.
        %Articles on electronic structure theory (mostly density functional theory) are highlighted in red.
        Data collected from the Web of Science on June 16th, 2021.
    }
    \label{tab:aps-most-cited}
\end{table*}

% WOS Query
% APS
% SO=(PHYSICAL REVIEW OR PHYSICAL REVIEW A OR PHYSICAL REVIEW A GENERAL PHYSICS OR PHYSICAL REVIEW ACCELERATORS "AND" BEAMS OR PHYSICAL REVIEW APPLIED OR PHYSICAL REVIEW B OR PHYSICAL REVIEW B SOLID STATE OR PHYSICAL REVIEW C OR PHYSICAL REVIEW D OR PHYSICAL REVIEW E OR PHYSICAL REVIEW FLUIDS OR PHYSICAL REVIEW LETTERS OR PHYSICAL REVIEW MATERIALS OR PHYSICAL REVIEW PHYSICS EDUCATION RESEARCH OR PHYSICAL REVIEW RESEARCH OR PHYSICAL REVIEW SPECIAL TOPICS ACCELERATORS "AND" BEAMS OR PHYSICAL REVIEW SPECIAL TOPICS PHYSICS EDUCATION RESEARCH OR PHYSICAL REVIEW X OR REVIEWS OF MODERN PHYSICS)

% ACS
% SO=(ACS APPLIED BIO MATERIALS OR ACS APPLIED ELECTRONIC MATERIALS OR ACS APPLIED ENERGY MATERIALS OR ACS APPLIED MATERIALS INTERFACES OR ACS APPLIED NANO MATERIALS OR ACS APPLIED POLYMER MATERIALS OR ACS BIOMATERIALS SCIENCE ENGINEERING OR ACS CATALYSIS OR ACS CENTRAL SCIENCE OR ACS CHEMICAL BIOLOGY OR ACS CHEMICAL HEALTH SAFETY OR ACS CHEMICAL NEUROSCIENCE OR ACS COMBINATORIAL SCIENCE OR ACS EARTH "AND" SPACE CHEMISTRY OR ACS ENERGY LETTERS OR ACS ENVIRONMENTAL SCIENCE "AND" TECHNOLOGY ENGINEERING OR ACS IEEE INTERNATIONAL CONFERENCE ON COMPUTER SYSTEMS "AND" APPLICATIONS PROCEEDINGS OR ACS INFECTIOUS DISEASES OR ACS MACRO LETTERS OR ACS MATERIALS LETTERS OR ACS MEDICINAL CHEMISTRY LETTERS OR ACS MISCELLANEOUS PUBLICATION OR ACS NANO OR ACS OMEGA OR ACS PHARMACOLOGY TRANSLATIONAL SCIENCE OR ACS PHOTONICS OR ACS SENSORS OR ACS SUSTAINABLE CHEMISTRY ENGINEERING OR ACS SYMPOSIUM SERIES OR ACS SYNTHETIC BIOLOGY OR BIOCHEMISTRY OR BIOCONJUGATE CHEMISTRY OR BIOMACROMOLECULES OR BIOTECHNOLOGY PROGRESS OR CHEMICAL RESEARCH IN TOXICOLOGY OR CHEMICAL REVIEWS OR CHEMISTRY OF MATERIALS OR CRYSTAL GROWTH DESIGN OR ENERGY FUELS OR ENVIRONMENTAL SCIENCE TECHNOLOGY OR ENVIRONMENTAL SCIENCE TECHNOLOGY LETTERS OR INDUSTRIAL ENGINEERING CHEMISTRY FUNDAMENTALS OR INDUSTRIAL ENGINEERING CHEMISTRY PROCESS DESIGN "AND" DEVELOPMENT OR INDUSTRIAL ENGINEERING CHEMISTRY PRODUCT RESEARCH "AND" DEVELOPMENT OR INDUSTRIAL ENGINEERING CHEMISTRY RESEARCH OR INORGANIC CHEMISTRY OR JACS AU OR JOURNAL OF AGRICULTURAL "AND" FOOD CHEMISTRY OR JOURNAL OF CHEMICAL EDUCATION OR JOURNAL OF CHEMICAL INFORMATION "AND" COMPUTER SCIENCES OR JOURNAL OF CHEMICAL INFORMATION "AND" MODELING OR JOURNAL OF CHEMICAL THEORY "AND" COMPUTATION OR JOURNAL OF MEDICINAL "AND" PHARMACEUTICAL CHEMISTRY OR JOURNAL OF MEDICINAL CHEMISTRY OR JOURNAL OF NATURAL PRODUCTS OR JOURNAL OF ORGANIC CHEMISTRY OR JOURNAL OF PHYSICAL CHEMISTRY OR JOURNAL OF PHYSICAL CHEMISTRY A OR JOURNAL OF PHYSICAL CHEMISTRY B OR JOURNAL OF PHYSICAL CHEMISTRY C OR JOURNAL OF PHYSICAL CHEMISTRY LETTERS OR JOURNAL OF PROTEOME RESEARCH OR JOURNAL OF THE AMERICAN CHEMICAL SOCIETY OR JOURNAL OF THE AMERICAN CHEMICAL SOCIETY FOR SPECTROMETRY)

Records of the pervasive use of these tools can be found in the scientific literature.
In a 2014 survey, van Noorden et al.\ found that 12 of the top 100 most cited papers of all time were on density-functional theory \cite{VanNoorden2014}.
As with other examples in van Noorden's list, the flood of citations are indicative of papers being cited by the many practitioners (here: of density-functional theory) rather than the few method developers.
%-- here: practitioners of density functional theory.
If one focuses the analysis on articles published in physics journals, the footprint of density-functional theory grows even further:
For example, table~\ref{tab:aps-most-cited} shows the top ten most cited papers published by journals of the American Physical Society.
All of them are related to density-functional theory and its application.

%In the chemistry field, the situation is less extreme (SI)

\begin{table*}
    \centering
    \begin{tabular}{llllS[table-format=7.0]} \toprule
        Software & Version & Main language& {Contributors 2020} & {Lines of code} \\ \midrule
        NWChem \cite{Apra2020}&7.0.2&Fortran& 11 &5034139\textsuperscript{*} \\
        LAMMPS \cite{Plimpton1995}&27May2021&C++& 51 & 1039872  \\
        CP2K \cite{Kuhne2020a}&8.2&Fortran & 38 & 911367 \\
        ABINIT \cite{Gonze2020}&9.4.2&Fortran& 21 & 719048 \\
Quantum ESPRESSO \cite{Giannozzi2020}&6.7.0&Fortran& 43& 604666 \\
GROMACS \cite{Abraham2015}&2021.2&C++& 25 & 589108 \\
Psi4 \cite{Smith2020}&1.3.2&C++& 34 & 496858 \\
SIESTA \cite{Garcia2020}&5.0.0-alpha&Fortran& 4 & 411829 \\
Octopus \cite{Tancogne-Dejean2020}&10.5&Fortran& 10 & 297825 \\
OpenMM \cite{Eastman2017}&7.5.1&C++& 21 & 256452 \\
xtb \cite{Bannwarth2021}&6.4.1&Fortran& 18 & 124344 \\\hline
Linux kernel&5.13&C& >4000 \cite{LinuxHistory2020} & 20904410\\
%\bottomrule
    \end{tabular}\\
    \caption{
        Counting lines of code for 11 popular open-source atomistic simulation engines (and the Linux kernel for comparison), using the latest releases as of June 2021.
        Line counts are determined by cloc v1.6.0 \cite{Danial2021} and exclude blank lines, comments, and markup languages (detailed reports in the supporting information).
        Contributors for the year 2020 were determined by counting the number of different committers to the source code from January 1st 2020 to January 1st 2021 (numbers for the Linux kernel are from 2019).
        \textsuperscript{*} Roughly 3 million lines of code of NWChem are computer-generated.
    }
    \label{tab:lines-of-code}
\end{table*}

There used to be a time when it was commonplace for computational condensed-matter physicists and quantum chemists to write their own electronic-structure code, and many of the atomistic-simulation engines that are in broad use today have started this way.
Over the years, however, many of these engines have developed into complex software distributions.
Table \ref{tab:lines-of-code} shows counts for the lines of code in some of today's popular open-source simulation engines: they range from hundreds of thousands to millions of lines of code, typically written in Fortran or C++, with similar numbers being reported for commercial packages \cite{Krylov2015}.
While statistics like these are by no means accurate measures of code complexity (and developers follow different approaches to packaging and outsourcing of functionality to external libraries), they nevertheless suggest that many of these code bases are too large to be sustained by any single person.
%As the compute power available to scientists has continued increasing, so has the complexity of the models that can be treated computationally. 
%This can range anywhere from treating very large systems (DFT on millions of atoms [cite])
%to very demanding methods (xx scaling to xxx cores).

This poses important questions for how to sustain these software projects going forward: questions of funding, business models, and software licenses.
Proponents of the open-source route argue that it democratizes research and education by removing barriers for both users and developers, and that science carried out with commercial software is harder to verify and reproduce \cite{Stodden2009,Gezelter2015,Lehtola2021}.
The open-source model can also be adopted irrespective of the size of a code's target user group, while commercial activity tends to require a minimal market size.
On the other hand, the strong focus on innovation and development often found in open-source scientific software can negatively impact usability and quality of documentation \cite{Swarts2019}.
% Thought: Since contributions to open-source software come from tech-savvy users who are trying to solve their own problem,
% documentation for novice users is easily neglected (and there is no driving fore for it).
Proponents of commercial licenses argue that making academic software accessible to a broader community is a technical task best left to professional software engineers, and that the resulting gain in scientific productivity can easily outweigh the license fees that pay for it \cite{Krylov2015}.
We note that open-source and commercial activity do not necessarily exclude each other: numerous software companies are built around an open-source core, and we start seeing first examples in atomistic modelling as well (e.g. Molcas/OpenMolcas \cite{Fdez.Galvan2019}).
Overall, it has been suggested that ``scientific publications are a more sound metric [of the scientific impact of software] than either the price of a product or whether its source code is available in the public domain'' \cite{Krylov2015}.
Providing a peek into this citation record is one of the reasons for creating the \atsoft collection.

%This is a question that the atomistic.software collection can help provide some insight into, at least as far as academic users are concerned (it is not tracking use in industry, where commerical software is more prevalent).

The other reason is a practical one: 
When young scientists start their first research project in atomistic simulations, they often have no grasp of the extent of this software ecosystem, let alone of current trends in the field -- at least this is the personal experience of the authors of this review.
Software choices in research groups are therefore often informed by what other members of the group already use.
This makes sense: colleagues have vetted the code for the type of problems the research group is working on, and built up expertise around which of the many knobs to turn in order to find the sweet spot between efficiency and accuracy.
%over the years, a zoo of different codes has developed.
%When one of the co-authors (TL) started his Ph.D.\ in 2010, he had no idea of the extent of this software ecosystem, let alone of current trends in the field, and one can assume the same holds true for many of those starting their career in the field today.
%While writing a section on best open-science practises for a review article, the question arose: given that in the field of computational materials science/solid state physics/ chemistry we are blessed with great open-source software that can do most of what the (equally great) commercial codes do, how are those two groups faring with respect to each other and can one identify trends in their usage?
%At that point I was aware of a couple of codes that had moved to more open licenses - e.g. CASTEP and Molcas - but I had no data at the ecosystem level, and that's when I started reaching out to Luca Ghiringelli at Nomad about building upon their list.

@@

%Today, there is a whole zoo of engines that makes it difficult to keep up even for experienced practitioners. 
%As a young scientist, you probably don't know where to start.

%Many young scientists today start using the engine that is already used by others in the group where they start their research project / master's degree / Ph.D.

@@

%The following sections introduce the \href{atomistic.software}{atomistic.software} collection aimed at helping researchers with some of these questions.

@@

%\begin{itemize}
%    \item introduce the collection (origins, content, scope)
%    \item discuss methods used for data collection\\
%        this may include 
%    \begin{itemize}
%        \item looking at examples of codes and comparing results of citations of a key paper vs the query-based approach
%        \item comparison google scholar counts vs web of science
%        \item comparison to related efforts in other fields (I'm not currently aware of any but will look \& ask around for them - suggestions welcome!)
%    \end{itemize}
%    \item point out limitations of the methods
%    \item at the same time, explain how to explore it through \url{atomistic.software}\\ 
%        %\begin{itemize}
%        %    \item how to nagivate the table (links to code homepages; classifications with explanation on hover; links to google scholar queries)
%        %    \item how to search / filter; select years; ...
%        %\end{itemize}
%        (this will involve a number of figures)
%\end{itemize}

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/overview}
    % picture taken with:
    % * incognito mode (to hide visited links)
    % * 125% magnification
    % * removed "tags" and "installation" column in order to reduce width
    % * keep sidebar unfolded (allows for more horizontal compression)
    % * on Retina display
    % * 8 codes in list for optimal height
    \caption{
        Overview table of atomistic simulation engines, sorted by how often they are referenced on Google Scholar during the previous year (here: 2020).
        A drop-down menu provides access to annual citation data reaching back to the year 2010.
    }
    \label{fig:overview}

@@

%TODO: Add comment on CAGR
%The compound annual citation growth rates reported on  are computed starting from the first year with 10 or more citations in order to be able to provide a growth vector also for new additions to the collection.
%We note that the concept of a growth rate implies exponential growth, which is not necessarily a good model to describe the data.
%An alternative approach would be to model linear growth in terms of additional citations per year.

%\begin{figure}
%    \includegraphics[width=0.5\textwidth]{figures/vasp}
%    \caption{
%        Citation trend for the VASP code.
%    }
%    \label{fig:vasp}
%\end{figure}

@@

\subsection{Trends}
%Discuss trends at the ecosystem level
%        (this will involve a number of figures)
%
%\begin{itemize}
%    \item commercial vs free
%    \item source code availability
%    \item (possibly) look at codes with large growth rates
%    \item (possibly) "the rise of the quantum materials science codes" (DFT codes with periodic boundary conditions seem to have grown a lot vs quantum chemistry \& force field codes)
%    \item (possibly) citation growth in atomistic software compared to general citation growth in STEM
%\end{itemize}

@@

% Could mention here some other aspect, e.g. wfm vs peridoic DFT
% or having a look at some of the quickly growing codes

@@

%There are a number of interesting outlooks that can be discussed here:
